---
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(plyr)
library(ggplot2)
```

#Loading and preprocessing the data

```{r}
df <- read.csv(unz("activity.zip", "activity.csv"), stringsAsFactors = FALSE)

head(df)

```


#What is mean total number of steps taken per day?

We begin by 

  1. Calculating the total number of steps taken per day
  
  2. Building a histogram of the total number of steps taken each day
  
  3. And calculating the mean and median of the total number of steps per day across all the days. 


```{r, warning = FALSE}

totalSteps <- ddply(df, c("date"), summarize, stepsPerDay = sum(steps, na.rm = TRUE))
ggplot(data = totalSteps, aes(stepsPerDay )) + geom_histogram()

```

```{r}
mean <- mean(totalSteps$stepsPerDay)
median <- median(totalSteps$stepsPerDay)
data.frame(mean, median)

```

#What is the average daily activity pattern?

To look into this we will 

  1. Make a time series plot of the average number of steps taken. This is the average by interval across days. 
  
  2. Identify the 5-minute interval which on average contains the maximum number of steps. 

```{r}
byInterval <- ddply(df, c("interval"), summarize, aveSteps = mean(steps, na.rm = TRUE))

ggplot(data = byInterval) +
    geom_line(aes(x=interval, y = aveSteps, group = 1)) +
    labs(x = "5 minute interval", y = "Average steps per interval") + 
    ggtitle("Average Daily Activity Pattern")

maxSteps <- max(byInterval$aveSteps)
interval<- byInterval[byInterval$aveSteps == max(byInterval$aveSteps),]$interval
data.frame(maxSteps, interval)

```

#Imputing missing values

1. First let's look at the number of missing values. 

```{r}
rowsWithNAs <- dim(is.na(df))[1]
rowsWithNAs
```


2. That's a lot. We can use a prebuilt method to fill these in using the "mice" library and the pmm (predictive mean matching) method. This function call generates 1 completed data frame with imputed values. 

3. We'll build a new dataset with the imputed values included. 

```{r}
library(mice)
tempdf <- mice(data = df, m = 1, method = "pmm", maxit = 10)
completedf <- complete(tempdf, 1)
```

4. Now let's look at a histogram of the total number of steps taken each day after missing values are imputed along with the new mean and median of average total steps taken across days. 

```{r}
completeTotalSteps <- ddply(completedf, c("date"), summarize, stepsPerDay = sum(steps))
ggplot(data = completeTotalSteps, aes(stepsPerDay)) + geom_histogram()

mean <- mean(completeTotalSteps$stepsPerDay)
median <- median(completeTotalSteps$stepsPerDay)
data.frame(mean, median)

```

We can see that we no longer have as many days with 0 total steps. And because we've filled in missing values, both our mean and median for total steps per day have increased. 

#Are there differences in activity patterns between weekdays and weekends?

To answer this question we first need to 

  1. Identify dates in the data as weekdays or weekends.
  
  2. Average steps per interval across all weekend days and again across all weekday days. 
  
  3. Plot the resulting averaged data in separate panels. 

```{r}
completedf$day <- weekdays(as.Date(completedf$date))
completedf$weekend <- ifelse(completedf$day %in% c("Saturday","Sunday"), "Weekend", "Weekday")

res <- ddply(completedf, c("interval","weekend"), summarize, aveSteps = mean(steps, na.rm = TRUE))

ggplot(res, aes(interval, aveSteps)) + geom_line() + facet_grid( .~ weekend)

```

Sure looks like there is an activity pattern difference! That early morning spike in movement during the week gets smoothed out and pushed back into a nice slow morning in the weekend. And the weekend shows higher activity levels in the middle of the day, hopefully because fewer people are sitting around at desks. 

